Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-07 method.tex
some rearrangment
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@948 ccdd05d5-56c1-461a-bf98-85f6876523b4

In this section we describe a decomposition of the \CBC{} waveform
space that reduces low-latency filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 joint science
runs.  We combine this with the \SVD\ rank-reduction method described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Equation~(\ref{eq:SNRTD}) can be implemented in \TD as an
\fir\ filter, requiring $\mathcal O(\numtmps \tmpsamps)$ floating point
operations per sample.  However, it is typically much more computationally
efficient to use the convolution theorem and the \fft\ to implement fast
convolution in the \FD, requiring only $\mathcal O(\numtmps
\lg \tmpsamps)$ operations per sample with increased latency.

\editorial{if we are gonna stick with lloid we should define it}
\subsection{The \lloid\ method}

Here we explore a method for reducing the
computational cost of a \TD\ search for compact binary coalescence.  We will
give a truly zero-latency algorithm that competes in terms of floating point
operations per second with the conventional \FD\ method, which requires a
signficant latency in order to be computationally cheap. Our method, \lloid{},
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.

The first transformation is to chop the \TD{} templates up into \emph{time
slices}.  Since each template slice is disjoint in time, the resulting set for a
single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim$100 samples
from
%
\editorial{Where do these numbers come from? CHAD: I have actually changed the statement to 100 for the early time slices.}
%
the early part of the waveform and allows the filters to be evaluated at about
$\sim$100 times lower rate.  This amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the singular value decomposition to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of $\sim$100.
The combined methods reduce the number of floating point operations to
the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}.  A numerical technique based on the singular value
decomposition (\SVD) of inspiral template banks is demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ is the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps.  In the next section we
describe how we form our novel detection statistic using the time-slice
decomposition and the \SVD.

@@

\Table{\label{table:flops}Computational cost in \flops\ of the \TD\ method, the \FD\ method, and \lloid\ for the bank of $2 \times 657$ templates sampled at 4096 Hz for chirp masses $1.1955 \leqslant \mathcal{M} / M_\odot \leqslant 1.2045$.  A block size of $D = 2 N$ is used for the \fft{}s in the \FD\ method.  Both the \FD\ method and \lloid\ are five orders of magnitude faster than the conventional \TD\ method.  However, the \FD\ method has a latency of over half of an hour, whereas \lloid\ has no latency at all.}
\br
Method & \flops\ & latency (s) \\
\mr
Time-domain (\TD) convolution & $2.4\times10^{13}$ & 0 \\
Frequency-domain (\FD) convolution & $2.6\times10^8$ & 2201 \\
\lloid\ (this work) & $4.7\times10^8$ & 0 \\
\br
\end{tabular}
\end{indented}
\end{table}

@@

Taking into account the decimation of the detector data, the orthogonal \fir\ filters, the reconstruction of the time-sliced templates, the interpolation of \SNR\ from previous time slices, and the accumulation of \SNR, in total \lloid\ requires
$$
\sum_{\mathclap{s=0}}^{\mathclap{S-1}} \left( 2 \slicessamps \numsvdtmps + 2 \numtmps \numsvdtmps + \numtmps \right) f^s + 2\sum_{\mathclap{f \in \{f^s\}}} \left( N^\shortdownarrow + \numtmps N^\shortuparrow \right) f \textrm{ \flops}\,.
$$
The second sum is carried out over the set of sample rates of all of the time slices rather than over the time slices themselves because some time slices may have the same sample rates.

\subsubsection{Extrapolation of the computational cost to an Advanced \LIGO{} search}
\editorial{Drew: I think this section be the last section of the results section. Should we move it to be Sec. 4.2? Nick: I'm inclined to agree with Drew, though I see that we want to make explicit that it's an analysis of the algorithm and not of the implementation and the current placement aids that. Also, the given FLOPS numbers correspond to what choice of \SVD{} tolerance and $N^\shortuparrow$?}

Table~\ref{table:flops} shows that the \lloid\ method requires $\sim 10^8$ \flops\
to cover a sub-bank of comprising $\sim 10^2$ out of $\sim 10^5$ sets of intrinsic
source parameters.  Given that modern (ca. 2011) workstations can sustain
computation rates up to $\sim 10^{10}$ \flops{}, and assuming that other regions of the
parameter space have similar computational scaling, an entire search could be implemented
with $\gtrsim 10$ machines.

By comparison, using the \TD\ method to achieve the same latency costs $\sim 10^{13}$ \flops\ for this particular sub-bank, and so it would require $\sim 10^6$ present-day machines.  Presently, the \LIGO{} Data Grid consists of only $\sim 10^4$ machines.

\begin{comment}

% Do we want to spend this much copy on the whitener?
% What matters is that we have picked a low-latency whitening procedure.

\subsection{Data Whitening}
Matched filtering the SVD basis vectors has motivated a decoupling of the
whitening routine from the matched filtering engine. This was necessary in
order weight templates appropriately by whitening them before the SVD
calculation, and, as such, we have also moved the data whitening outside the
match filter engine. We have chosen to whiten the data using a running
geometric average of the PSD computed by Hann windowing 8 second buffers of
data and using 50\% overlapping buffers. The running geometric average is
updated once per buffer from a median of the recent PSDs. We find that this
algorithm is extremely fast to converge to an accurate PSD and also very robust
against glitches.

An inspiral, to leading order, has gravitational-wave frequency evolution given
by
%
\begin{equation}
%
f(t) = \frac{1}{8 \pi \mathcal{M}} \left( \frac{t_c - t}{5 \mathcal{M}}
\right)^{-3/8}\;,
%
\end{equation}
%
whose time derivative is given by
%
\begin{equation}
%
\frac{df}{dt} = \frac{3}{320 \pi \mathcal{M}^2} \left( \frac{t_c - t}{5
\mathcal{M}} \right)^{-11/8}\;.
%
\end{equation}
%
Combining these, we find the frequency evolution as a function of frequency to
be
%
\begin{equation}
%
\frac{df}{dt}(f_0) = \frac{3}{320 \pi \mathcal{M}^2} \left( 8 \pi \mathcal{M}
f_0 \right)^{3/11}\;,
%
\end{equation}
%
which can be inverted to get the minimum frequency which has a given frequency
derivative,
%
\begin{equation}
%
f_0 = \frac{1}{8 \pi \mathcal{M}} \left( \frac{320 \pi \mathcal{M}^2}{3}
\frac{df}{dt} \right)^{11/3}\;.
%
\end{equation}

If we allow frequency bins to be affected by an injection only once within the
median history, we need to calculate the corresponding $df/dt$ for our PSD
calculation. An FFT of a buffer with length $T$ will have a frequency
resolution $df = 1/(2T)$. Hann windowing the data and overlapping buffers by
50\% introduces correlations in neighboring frequency bins of the FFT. This
means that we actually want the frequency to change by $df = 3/(2T)$ before the
next PSD calculation, which happens a $dt = T/2$ later, resulting in a minimum
$df/dt = 3/T^2$.

Combining these results we find the lower frequency bound for our injections,
assuming a chirp mass for a binary with $m_1 = m_2 = 1 M_{\odot}$ and $T = 8$
s, is 23 Hz.

We want to compute the equivalent injection density which would be biased as
much as lalapps\_inspiral's PSD estimator, which allows $\sim3$ injections per
2048 seconds. It computes the PSD by breaking the segment up into 16 256 second
chunks. These chunks are then combined into to sets of 8 chunks each. The
median of each set is then averaged across the sets for each frequency bin to
produce the PSD estimate for that 2048 second segment. This procedure results
in an average injection density of 1.5 per 8 PSDs in the median history, which
is comparable to LLOID's 1 per 7 PSDs in the median history. Since LLOID has an
injection history of 7 PSDs which span 32 seconds, this means LLOID can perform
one injection every 32 seconds, or 64 injections per 2048 second segment, more
than an order of magnitude increase in density over lalapps\_inspiral.
\end{comment}

>>>>

In this section we describe a decomposition of the \CBC{} waveform
space that reduces time-domain (\TD) filtering cost sufficiently to allow for the
possibility of \earlywarning\ detection with modest computing requirements.  We
expand on the ideas of~\cite{Marion2004, Buskulic2010} that describe a
multiband decomposition of the compact binary parameter space that resulted in
a search with minutes latency in \LIGO{}'s S6 and Virgo's VSR3 joint science
runs.  We combine this with the \SVD\ rank-reduction method described
in~\cite{Cannon:2010p10398} that exploits the redundancy of the template banks.

@@

Equation~(\ref{eq:SNRTD}) can be implemented in the time-domain as an
\fir\ filter, requiring $\mathcal O(\numtmps \tmpsamps)$ floating point
operations per sample.  However, it is typically much more computationally
efficient to use the convolution theorem and the \fft\ to implement fast
convolution in the frequency-domain, requiring only $\mathcal O(\numtmps
\lg \tmpsamps)$ operations per sample with increased latency.

\subsection{The \lloid\ method}

Here we explore a method for reducing the computational cost of a \TD\ search
for compact binary coalescence.  We will give a truly zero-latency algorithm
that competes in terms of floating point operations per second with the
conventional \FD\ method, which requires a signficant latency in order to be
computationally cheap. Our method, \lloid\footnote{\lloid, The term strictly
means \emph{Low Latency Online Inspiral Detection} and was a code name for
this procedure as it was being developed. It is now known in some GW circles.
We feel that although the name doesn't carry much specific information about
the details of the algorithm it is far too late to do anything about it now.}
involves two transformations of the template waveforms that produce a set of
orthogonal filters with far fewer coefficients than the original templates.

The first transformation is to chop the \TD{} templates up into \emph{time
slices}.  Since each template slice is disjoint in time, the resulting set for
a single template is orthogonal.  Given the chirp-like structure of the
templates, the early time slices have significantly lower bandwidth and can be
safely downsampled.  The downsampling removes a factor of $\sim$100 samples
from the early part of the waveform and allows the filters to be evaluated at
about $\sim$100 times lower rate.  This amounts to more than a factor of 10000
reduction in the floating-point operations per second required to filter the
early part of the waveform.  However, the resulting filters are still not
orthogonal across the mass parameter space, and are in fact highly redundant.
We use the singular value decomposition to produce an orthogonal filter set
from the time-sliced templates~\cite{Cannon:2010p10398}.  We find that this
reduces the number of sample points of the filters by another factor of
$\sim$100.  The combined methods reduce the number of floating point operations
to the level where they are competitive with the conventional high-latency \FD\
matched filter approach.  In the remainder of this section we describe the
\lloid\ algorithm in detail and provide some basic computational cost scaling.

@@

As described previously, the template banks are, by design, highly correlated.
It is possible to greatly reduce the number of filters required to achieve a
particular minimum match by designing an appropriate set of \SVD\ {\em
basis templates}.  A numerical technique based on the singular value
decomposition (\SVD) of inspiral template banks is demonstrated
in~\cite{Cannon:2010p10398}.  Similarly, the time sliced templates described
above can be approximated to arbitrary accuracy by expansion into a set of
\SVD\ \emph{basis templates}, $u_l^s[k]$, related to the original time
sliced templates through the \emph{reconstruction matrix},
$v_{il}^s\sigma_l^s$:
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
%
\editorial{Say a word or two about \SVD\ tolerance?}%
%
The parameter $\numsvdtmps$ is the number of basis templates that are kept in
the approximation.  This determines the \SVD\ tolerance, which affects the
\SNR\ loss due to the approximation.  The authors of \cite{Cannon:2010p10398}
showed that high accuracy could be achieved with far fewer basis templates than
templates in the original template bank.  We find that when combined with the
time-slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps\ and improves on
\cite{Cannon:2010p10398} by nearly an order of magnitude.  In the next section
we describe how we form our novel detection statistic using the time-slice
decomposition and the \SVD.

@@

@@

Taking into account the decimation of the detector data, the orthogonal \fir\ filters, the reconstruction of the time-sliced templates, the interpolation of \SNR\ from previous time slices, and the accumulation of \SNR, in total the \lloid\ algorithm requires
$$
\sum_{\mathclap{s=0}}^{\mathclap{S-1}} \left( 2 \slicessamps \numsvdtmps + 2 \numtmps \numsvdtmps + \numtmps \right) f^s + 2\sum_{\mathclap{f \in \{f^s\}}} \left( N^\shortdownarrow + \numtmps N^\shortuparrow \right) f \textrm{ \flops}\,.
$$
The second sum is carried out over the set of sample rates of all of the time slices rather than over the time slices themselves because some time slices may have the same sample rates.
